# Lecture

## Machine epsilon
The first exercise from this lecture was to write a program to determine the relative machine epsilon.

```{r}
# First we define a function that takes no input
machine.eps <- function() {
  # We guess that the machine epsilon is less than 1, so this is where we start
  eps <- 1
  # Then we want to add eps times one half and check if the result is neq 1
  while (1 + eps / 2 != 1) {
  # If that is not the case we divide eps by 2 and check the condtion again
  eps <- eps / 2
  }
  # When the condition no longer hold our result is the relative machine
  # epsilon, i.e. the point where the computer can no longer distinguish 1 and 1 + eps
  return(eps)
}
machine.eps()
```

## Finding roots with search algorithms

### Defining the objective function
Then we defined a function.

```{r}
# The function is first defined
f <- function(x) {x^3 - x^2 - 1}
```

Tested it for relevant values

```{r}
f(0)
f(2)
f(1.5)
```

And plotted the output for the function

```{r}
a <- 0
b <- 2
curve(f, a, b)
abline(h = 0, col = "gray")
```

We then went about finding the root of this function, first by naive search and thereafter using the bisectional method.

### Naive search

```{r}
set.seed(1)

x_candidate <- runif(n = 1, a, b)
x_candidate

f_candidate <- f(x_candidate)
f_candidate

f_tolerance <- 1e-5

naive_search <- function() {
    iterations <- 0
  while (abs(f_candidate) > f_tolerance) {
    x_candidate <- runif(n = 1, a, b)
    f_candidate <- f(x_candidate)
    iterations <- iterations + 1
  }
  cat(
    "x candidate:\t", 
    x_candidate, "\t", 
    "f candidate:\t", 
    f_candidate, "\t", 
    "number of iterations:\t", 
    iterations)
}
naive_search()
```

### Bisectional method

```{r}
bisection <- function(f, a, b, tolerance = 1e-5, max_iterations = 1000) {
  iterations <- 0
  for (i in 1:max_iterations) {
    c <- (a + b)/2
    if (sign(f(a)) == sign(f(c))) {
      a <- c
    } else {
      b <- c
    }
    iterations <- iterations + 1
    if (abs(f(c)) < tolerance){
      break
    }
  }
  cat("Root:\t", c, "f candidate:\t", f(c), "In:\t", iterations, "iterations")
}
bisection(f, 0, 2)
```

