# Lecture

The exercises for this lecture revolves around line search methods.
The first exercise is to implement steepest descent.

## Steepest descent

```{r}
# First we need a function that we want to minimize
obj_f <- function (x) {x^2 + 2*x + 3}
# We also need the gradient of the function
obj_g <- function(x) {2*x + 2}

# Then we can implement steepest descent for this function
steepest_descent <- 
  function (x_0, 
            alpha = 0.1, 
            f = obj_f, 
            g = obj_g, 
            max_it = 100) {
  #First we plot the function in an arbitrary interval
  curve(f, -3, 4)
  # Insert the initial point in the plot
  points(x_0, f(x_0))
  # And set our x_k equal to the x_0
  x_k <- x_0
  for (i in 1:max_it) {
    # Then we find the steepest descent direction
    p_k <- -g(x_k)
    # We take a step of length alpha in the direction just found
    x_k <- x_k + alpha * p_k
    # Evaluate the function at this x_k
    f_k <- f(x_k)
    # And plot our new position
    points(x_k, f_k)
  }
  cat('f(x):',f_k,'\t','x:',x_k)
}

#The steepest descent implementation with x_0 = 4
steepest_descent(4)
```

## Backtracking line search
The nexto topic is to implement a backtracking line search algorithm for determining a more optimal step length, and thereafter include it in the steepest descent algorithm.

```{r}
# Our algorithm will take an initial position,
# objective function, the gradient of this
# function, rho (how much smaller to make the step),
# c (our tolerance) and alpha_hat (initial step length)
backtracking_line_search <- 
  function (x_0, 
            f = obj_f, 
            g = obj_g, 
            rho = 0.5, 
            c = 10e-4, 
            alpha_hat = 0.1) {
  # we set x_k equal to the initial position
  x_k <- x_0
  # we find the direction of steepest descent
  p_k <- -g(x_k)
  # the initial step length is set to alpha_hat
  alpha <- alpha_hat
  while (
    f(x_k + alpha * p_k) > 
    f(x_k + c * alpha * g(x_k) * p_k)
        ) {
    # If the condition is not met, we shrink the step
    alpha <- alpha * rho
  }
  alpha
}
backtracking_line_search(1)
```

### Combining steepest descent and backtracking
Now its time to include the backtracking in the steepest descent algorithm.

```{r}
steepest_descent_w_backtrack <- 
  function (x_0, 
            alpha = 1, 
            f = obj_f, 
            g = obj_g, 
            max_it = 100) {
  curve(f, -6, 3)
  points(x_0, f(x_0))
  x_k <- x_0
  for (i in 1:max_it) {
    # the step length is recalculated in 
    # each iteration of the for loop
    alpha <- backtracking_line_search(x_k)
    p_k <- -obj_g(x_k)
    x_k <- x_k + alpha * p_k
    f_k <- obj_f(x_k)
    points(x_k, f_k)
  }
}
steepest_descent_w_backtrack(-5)
```